### 2015-03-02 Manuscript/Missing_mammals-draft.tex
Full draft manuscript

\subsection{Data availability}
We extracted 1422 living mammal OTUs from the 256 matrices with a minimum of 6 characters and a maximum of 4541 per OTU. After removing all the matrices with less than 100 morphological characters, the number of extracted living mammals OTUs was down to 815. 11/28 orders have less than 25\% of taxa with morphological data at a species level and 24/28 orders have less than 75\% taxa with available morphological data. At the Genus level however only 3/28 orders have less than 25\% of taxa with morphological data and 16/28 have less than 75\%. Finally, at the family level no order has less than 25\% taxa with available morphological data and only 5/28 have less than 75\% (table~\ref{Table_morpho_taxa_proportion}).

@@

\subsection{Available data structure}
Among the orders containing at least one OTU with no available morphological data, only four orders are significantly clustered: Dirpotodontia, Carnivora and Chiroptera both at the species and the genus level and Rodentia at the species level only (table~\ref{Table_data_structure}).

@@

Two contrasted results are shown on figure ~\ref{Figure_example_coverageA} and ~\ref{Figure_example_coverageB} with randomly distributed data in Cetartiodactyla (figure ~\ref{Figure_example_coverageA}) and clustered available data in Carnivora (mainly Canidae; figure ~\ref{Figure_example_coverageB}).

@@

\section{Discussion}
Our results show that even though relations among living mammals are well known in a molecular framework (e.g. \cite{FritzTree,meredithimpacts2011,May-Collado-PeerJ}) only little morphological data is available. This is primarily due to the fact that that morphological studies are mainly focusing on fossil taxa rather than living taxa (e.g. \cite{O'Leary08022013,ni2013oldest}). However, data availability varies in its structure as well as across taxonomic levels. Higher taxonomic levels are always better sampled than lower ones (table~\ref{Table_morpho_taxa_proportion}) and within these taxonomic levels, the structure of available data is mostly random apart for four orders (table\ref{Table_data_structure}).

%Data collection
We know we might have missed some data but we performed a thorough and repeatable search and still managed to collected 1422 OTUs which represents 38\% of all living mammals

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{../andrew/citations.jpg}
\caption{ }
\label{bla}
\end{figure}

%\subsection{Data availability}
The amount of living taxa with no available morphological data was surprisingly high at the species level (24/28 orders have less than 75\% taxa with available morphological data) which can lead to topological misplacements of fossils in Total Evidence trees. The effect of missing living taxa with morphological data has no significant impact on inferring the correct topology up to 25\% of missing data, beyond that threshold, both the position of "flying" taxa and the clades are significantly different from the correct topology \cite{GuillermeCooper}. At the species level within mammals, it is therefore likely to see topological artefacts for the placement of fossil taxa. However, the data availability is more promising at higher taxonomic level (i.e. the Genus and the Family level). This value is probably more important to consider since there is a slight discrepancy between neontological species and paleontological species. While the second ones are described based on accurately measured morphology, genetic distance, spacial distribution and even behaviour, the second one are usually estimate based on morphological, spatial and temporal differences only. Because of the lack of behavioural and genetic data, it is likely that cryptic species are often ignored in palaeontological data. Therefore, most palaeontological studies are using the Genus level for their smallest taxonomic OTU. In this frame of palaeontological data usage, the data availablity at the Genus level in living mammals should be our priority concern.

%\subsection{Available data structure}
However, it is encouraging that for most orders (except Carnivora and Chiroptera), the few available data is randomly distributed, therefore the addition of fossil taxa is likely to not be biased by oversampling but will probably branch near the taxa that has the most similar available data (as expected). When only few data is available, the ideal scenario will be that the data is over-dispersed (i.e. that there is data in at least every sub-clade) in order to maximize the possibilities of the fossil to branch in the right clade. In the second scenario, when the data is randomly distributed we expected no special bias in the placement of the fossil \cite{GuillermeCooper}. However, in the third scenario, when the data is clustered we expect two major biases to occur: (1) first, the fossil will not be able to branch within a clade containing no data (e.g. Herpestidae in Carnivora; figure ~\ref{Figure_example_coverageB}), and (2) second, the fossil will have a higher probability, at random, to branch within the clade containing most of the available data (e.g. Canidae in Carnivora; figure ~\ref{Figure_example_coverageB}).

@@

%Caveats
In this study we only used the raw number of characters and not there similarity. This might not work in reality since the data may actually not overlap. It might be that actually the amount of available data here is lower due to non overlap of characters....

%Let's code some data
Since the only way to fix missing data issues is to collect the missing data, it queit encouraging to know that the morphological data is available through natural history museum collections and that online plateforms and tools exist and are opperational to facilitate this giant task through collaboration and sharing open source and traceable data!

>>>>

\subsection{Data availability}
We extracted 1422 living mammal OTUs from the 256 matrices with a minimum of 6 characters and a maximum of 4541 per OTU. After removing all the matrices with less than 100 morphological characters, the number of extracted living mammals OTUs was down to 815. 11/28 orders have less than 25\% of taxa with morphological data at a species level and 24/28 orders have less than 75\% taxa with available morphological data. At the Genus level however only 3/28 orders have less than 25\% of taxa with morphological data and 16/28 have less than 75\%. Finally, at the family level no order has less than 25\% taxa with available morphological data and only 5/28 have less than 75\% (table \ref{Table_morpho_taxa_proportion}).

@@

\subsection{Available data structure}
Among the orders containing at least one OTU with no available morphological data, only two orders are significantly clustered: Carnivora and Chiroptera both at the species and the genus level (table~\ref{Table_data_structure}).

@@

Two contrasted results are shown on figures \ref{Figure_example_coverageA} and \ref{Figure_example_coverageB} with randomly distributed data in Cetartiodactyla (figure \ref{Figure_example_coverageA}) and clustered available data in Carnivora (mainly Canidae; figure \ref{Figure_example_coverageB}).

@@

\section{Discussion}
Our results show that even though relations among living mammals are well known in a molecular framework (e.g. \cite{FritzTree,meredithimpacts2011,May-Collado-PeerJ}) only little morphological data is available. This is primarily due to the fact that that morphological studies are mainly focusing on fossil taxa rather than living taxa (e.g. \cite{O'Leary08022013,ni2013oldest}). However, data availability varies in its structure as well as across taxonomic levels. Higher taxonomic levels are always better sampled than lower ones (table \ref{Table_morpho_taxa_proportion}) and within these taxonomic levels, the structure of available data is mostly random apart for two orders (table \ref{Table_data_structure}).

%\subsection{Data availability}
The amount of living taxa with no available morphological data was surprisingly high at the species level: 24/28 orders have less than 75\% taxa with available morphological data (and two of the 28 orders are monospecific). This threshold of 75\% available data is the maximum amount of possible missing data (25\%) before having a significant effect on the topology of total evidence trees \cite{GuillermeCooper}. Beyond this threshold, there is a noteworthy displacement of the wildcard taxa (\textit{sensu} \cite{kearneyfragmentary2002}) as well as a important decrease in the conservation of clades \cite{GuillermeCooper}. At the species level within most orders of mammals, it is therefore likely to see topological artefacts for the placement of fossil taxa. However, the data availability seems to be less an issue at higher taxonomic level (i.e. the Genus and the Family level). This point is important to consider from a practical point of view because of the slight discrepancy between neontological and palaeontological species. While neontological species are described on accurately measured morphology, genetic distance, spacial distribution and even behaviour (e.g. \cite{kellymolecular2014}); palaeontological data can be based only on morphological, spatial and temporal data (e.g. \cite{ni2013oldest}). Because of this difference between neontological species (\textit{sensu} reproductive isolates) and paleontological species (\textit{sensu} morpho-species), most palaeontological studies are using the Genus level for their smallest taxonomic OTU (e.g. \cite{ni2013oldest,O'Leary08022013}). In this frame of palaeontological data usage, the data availability at the Genus level in living mammals should be our priority concern.

%\subsection{Available data structure}
Regardless the low level of available data, it is encouraging that for most orders (except Carnivora and Chiroptera), it is randomly distributed across the phylogeny. Therefore, it is likely that the addition of fossil taxa within these orders will not be biased by oversampling but will probably branch near the taxa that has the most similar available data (as expected). When only few data is available, the ideal scenario will be that the data is over-dispersed (i.e. that there is data in at least every sub-clade) in order to maximize the possibilities of the fossil to branch in the right clade. In the second scenario, when the data is randomly distributed we expected no special bias in the placement of the fossil \cite{GuillermeCooper}. However, in the third scenario, when the data is clustered we expect two major biases to occur: (1) first, the fossil will not be able to branch within a clade containing no data (e.g. Herpestidae in Carnivora; figure \ref{Figure_missing_data_problem} and \ref{Figure_example_coverageB}), and (2) second, the fossil will have a higher probability, at random, to branch within the clade containing most of the available data (e.g. a Carnivora fossil will have more chance to branch, randomly, to the Canidae clade than any other clade in Carnivora; figure \ref{Figure_example_coverageB}).

@@

%Caveats
In this study, we treated all morphological matrices to be equal in a similar way that molecular matrices are: for example if a matrix A contains 100 characters for taxa X and Y, and a matrix B contains 50 characters for taxa X and Z, we assumed that both matrices can be combined in a supermatrix way leading to a matrix containing 150 characters for taxon X, 100 for taxon Y and 50 for taxon Z. However, it is clear that morphological data cannot be treated that way \cite{Brazeau2011}. In fact, for the same taxon, some characters might overlap (e.g. matrix A has a character coding for the shape of a particular morphological feature and matrix B as a first character coding for the presence of this morphological feature and a second one coding for its shape; in this case, these three characters are compound characters \cite{Brazeau2011}). However, in reasonably sized matrices ($>$ 100 characters \cite{GuillermeCooper,harrisonamong-character2014}) it is more likely that a number of characters are consistently conserved among the different matrices (e.g. \cite{ross1998phylogenetic,seiffert2003fossil,marivaux2005anthropoid,seiffert2005basal,bloch2007new,kay2008anatomy,silcox2008biogeographic,seiffert2009convergent,tabuce2009anthropoid,boyer2010astragalar,seiffert2010fossil,marivaux2013djebelemur,ni2013oldest}). A conservative approach to avoid compound characters could be to select only the latest matrix for each taxon.

%Let's code some data
Following the recommendations in \cite{GuillermeCooper}, one should code morphological characters for a maximum of living species in order to improve the topology of total evidence trees containing both living and fossil taxa. Since the data for living mammals is usually easily available in world wide vast natural history collections, we propose that an increased effort in coding morphological characters from living species should be done via engaging in collaborative data collection projects through web portals such as \textit{morphobank} \cite{morphobank}.
